Deep mutational scanning of hemagglutinin helps distinguish the evolutionary fates of human H3N2 influenza virus lineages

### 2018-01-19 paper/manuscript.tex
added skeleton for post-Perth/2009 tree

@@

Our experiments were performed on the Perth/2009 HA, but we are scoring mutations on a phylogenetic tree that extends from 1968 to 2017.
Because the effects of mutations can vary with genetic background~\cite{gong2013stability,natarajan2013epistasis,harms2014historical,starr2016epistasis,starr2017alternative}, it is possible that the effects of mutations that we measured in the Perth/2009 HA might have shifted somewhat in other related HAs.
To explore this question, we scored the complete HA sequence $\mathbf{s}$ of every node in the phylogenetic tree by quantifying its average per-site sequence preference as 
\begin{equation}
F\left(\mathbf{s}\right) = \frac{1}{L}\displaystyle\sum_{r=1}^L \ln \pi_{r, s_r},
\end{equation}
where $\pi_{r, s_r}$ is the preference measured in our experiments (e.g., Figure~\ref{fig:logoplot}) for the amino acid $s_r$ at site $r$ in HA sequence $\mathbf{s}$, and $L$ is the length of the sequence.
Figure~\ref{fig:trunkvssidebranch}C shows that the sequences of trunk nodes tend to have higher preference than those of side branch nodes across the entire timespan, consistent with the finding that trunk mutations are generally more favorable than side branch mutations. 
However, the average per-site sequence preference increases as the nodes approach the Perth/2009 strain.
The Perth/2009 itself has among the highest per-site sequence preference of the entire tree, despite falling on a side branch.
These results suggest that there is some HA background-dependence in the effects of mutations, but that despite this fact our experimental measurements consistently reveal the selective advantage of the trunk over more than a half-century of viral evolution.

@@

@@

\subsubsection*{SI Tables}

Supply Word, RTF, or LaTeX files (LaTeX files must be accompanied by a PDF with the same file name for visual reference); include only one table per file. Do not use tabs or spaces to separate columns in Word tables.

\subsubsection*{SI Datasets}

>>>>

\setcounter{topnumber}{10}
\setcounter{bottomnumber}{10}
\setcounter{totalnumber}{10}

@@

Our experiments were performed on the Perth/2009 HA, but we are scoring mutations on a phylogenetic tree that extends from 1968 to 2017.
Because the effects of mutations can vary with genetic background~\cite{gong2013stability,natarajan2013epistasis,harms2014historical,starr2016epistasis,starr2017alternative}, it is possible that the effects of mutations that we measured in the Perth/2009 HA might have shifted somewhat in other related HAs.
To explore this question, we scored the complete HA sequence $\mathbf{s}$ of every node in the phylogenetic tree by quantifying its average per-site sequence preference as 
\begin{equation}
F\left(\mathbf{s}\right) = \frac{1}{L}\displaystyle\sum_{r=1}^L \ln \pi_{r, s_r},
\end{equation}
where $\pi_{r, s_r}$ is the preference measured in our experiments (e.g., Figure~\ref{fig:logoplot}) for the amino acid $s_r$ at site $r$ in HA sequence $\mathbf{s}$, and $L$ is the length of the sequence.
Figure~\ref{fig:trunkvssidebranch}C shows that the sequences of trunk nodes tend to have higher preference than those of side branch nodes across the entire timespan, consistent with the finding that trunk mutations are generally more favorable than side branch mutations. 
However, the average per-site sequence preference increases as the nodes approach the Perth/2009 strain.
The Perth/2009 itself has among the highest per-site sequence preference of the entire tree, despite falling on a side branch.
We do not think that this result means that Perth/2009 is actually higher fitness than the other strains.
Rather, if there are modest shifts in the effects of mutations as HA evolves, then sequences more distant from Perth/2009 could be scored as having lower preference simply because these shifts start to degrade the accuracy of our measurements.
But importantly, our experiments consistently show the trunk to be more favorable than the side branches regardless of whether we look at sequences that precede Perth/2009 (most of Figure~\ref{fig:trunkvssidebranch}C) or sequences that occur after Perth/2009 has diverged from the trunk (Figure~\ref{suppfig:future}). 
Therefore, our experimental measurements consistently reveal the selective advantage of the trunk over more than a half-century of viral evolution.

@@

\begin{suppfigure}
\comment{add figure}
\caption{\label{suppfig:future}
Our experiments show a selective advantage for the trunk even if we limit the analysis to sequences that occur after the split of Perth/2009 off the trunk.
(A) Phylogenetic tree showing all nodes that branch from the trunk before or contemporaneously with Perth/2009 (orange), all nodes that occur after the branch to Perth/2009 for which we can resolve the trunk (red) or side branches (blue), and all nodes for which it is not yet clear which sequences will be on the trunk or side branches.
In this figure, we restrict the analysis to the nodes in red or blue.
(B)  
Average per-site sequence preference (similar to Figure~\ref{fig:trunkvssidebranch}C) for all trunk and side-branch nodes that occur after the branch to Perth/2009.
(C)
The sequence preferences of nodes on the trunk generally exceed those of nodes on the side branches along the post-Perth/2009 portion of the tree.
Note that because there are relatively few post-Perth/2009 sequences for which we can currently distinguish trunk from side branch, the number of points in this plot is small, precluding meaningful statistical analyses of the type performed in Figure~\ref{fig:trunkvssidebranch}B.
}
\end{suppfigure}

@@

\clearpage

\subsubsection*{SI Datasets} 
\FloatBarrier
